---
title: "Chickens!"
author: "Teague O'Mara"
date: "11/11/2021"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Bring in some chicken GPS data

```{r}
library(move)
library(tidyverse)

cluck <- move("./data/Chickens Gallus gallus domesticus Hammond_GPS.csv")
cluck$birdID <- cluck@trackId
unique(cluck$birdID)

```

We have the locations from our 5 birds. We could generate our usual movement metrics from them (speed, turning angle, etc). 


```{r}
cluck$speed <- unlist(lapply(speed(cluck), c, NA))
cluck$angle <- unlist(lapply(angle(cluck), c, NA))

ggplot(cluck@data)+
  geom_density(aes(x = speed, fill = birdID), alpha = 0.5)+
  scale_fill_viridis_d()+
  xlim(0, 2.5)+
  xlab(expression(paste("speed m", s^-1)))
```




But Beachy wants to know if his chickens have differential social relationships. To do this, we'll create a chicken social network to see who the outcasts are.

```{r}
library(spatsoc)
library(asnipe)
library(data.table)
library(igraph)

#First we need to turn our move object into a data table.
cluck_dt <- as.data.table(cluck)

#Group our timstamps into 1 minute bins
group_times(cluck_dt, datetime = "timestamp", threshold = "1 minute")

#Generate our spatial network 
group_pts(cluck_dt, 
          threshold = 5, 
          id = "birdID", 
          coords = c("utm.easting", "utm.northing"), 
          timegroup = "timegroup")

gbi <- get_gbi(cluck_dt, group = "group", id = "birdID")

net <- get_network(gbi, data_format = "GBI", association_index = "SRI")

## Plot the spatial graph
g <- graph.adjacency(net, 'undirected',
                     diag = FALSE, weighted = TRUE)
plot(g)

```

Gardinea's GPS tag stopped working early, so it wouldn't be surprising if she falls outside of the main network. We can also see if there are any communities in our network. A community is just a group of individuals that asssociate more than others. We can also extract out metrics like network strength & centrality.

```{r}
observedNetworkStats <- data.table(
  centrality = evcent(g)$vector,
  strength = graph.strength(g),
  degree = degree(g),
  ID = names(degree(g))
)

coms <- cluster_fast_greedy(g)
plot(g, mark.groups = coms)

```

Let's look at some acceleration data.

```{r}
##ACC data
options(digits.secs=4)
library(lubridate)

acc <- read.csv("./data/Chickens Gallus gallus domesticus Hammond_ACC.csv")
acc$timestamp <- as.POSIXct(acc$timestamp, format = "%Y-%m-%d %H:%M:%OS", tz = "UTC")

#Need to convert the bursts to rows
acc_L <- split(acc, f = acc$event.id)
accLong <- lapply(acc_L, function(g){
  accAxesL <- lapply(str_split(as.character(g$accelerations.raw),' '), as.numeric) #get just the acc
  accAxesL_v <- unlist(accAxesL) #turn the acc values into a numeric vector
  g$TotalNumberSamples_rec <- as.numeric(length(accAxesL_v)) #how many samples are there?
  sampfreq <- g$acceleration.sampling.frequency.per.axis
  
  #Now take every 3rd observation for each axis
  x <- accAxesL_v[seq(1,g$TotalNumberSamples_rec,3)]
  y <- accAxesL_v[seq(2,g$TotalNumberSamples_rec,3)]
  z <- accAxesL_v[seq(3,g$TotalNumberSamples_rec,3)]
  timestamp <-  rep(g$timestamp, length(x))
  for(i in 1:(length(timestamp)-1)){
    j = i+1
    timestamp[j] = timestamp[i]+1/sampfreq
  }
  
  tmp_df <- bind_cols(g$event.id,
                      g$individual.local.identifier,
                      g$tag.local.identifier,
                      g$acceleration.sampling.frequency.per.axis,
                      g$TotalNumberSamples_rec,
                      timestamp,
                      x,
                      y,
                      z)
  names(tmp_df) <- c("event.id", "individual.local.identifier", "tag.local.identifier", "acceleration.sampling.frequency.per.axis", "TotalNumberSamples_rec", "timestamp", "x", "y", "z")
  return(tmp_df)
}) 

acc_df <- do.call("rbind", accLong)

#Create places for new variables
acc_df$odba <- NA
acc_df$staticX <- NA
acc_df$staticY <- NA
acc_df$staticZ <- NA
acc_df$pitch <- NA

acc_df_L <- split(acc_df, f = acc_df$individual.local.identifier)

#Define window length
sampfreq <- 16
nSecs <- 5
interval <- sampfreq*nSecs

library(zoo)
odbaL <- lapply(acc_df_L, function(temp){
  print("Calculating static accelerations and pitch")
   # Calculate mean acceleration over a moving window
  temp$staticX <- rollmean(temp$x, interval, fill = NA)
  temp$staticY <- rollmean(temp$y, interval, fill = NA)
  temp$staticZ <- rollmean(temp$z, interval, fill = NA)
  
  # Calculate pitch based on mean acceleration in each axis
  temp$pitch <- atan(temp$staticX/(sqrt((temp$staticY^2)+(temp$staticZ^2))))*(180/pi)
  
  # Calculate dynamic acceleration
  print("Calculating dynamic accleration")
  temp$dynamicX <- temp$x - temp$staticX
  temp$dynamicY <- temp$y - temp$staticY
  temp$dynamicZ <- temp$z - temp$staticZ
  temp$ODBA <- abs(temp$dynamicX) + abs(temp$dynamicY) + abs(temp$dynamicZ)
  return(temp)
})

odba_df <- do.call("rbind", odbaL)

```


Now we have measures of odbs & pitch for our birds.


```{r}
ggplot(odba_df)+
  geom_histogram(aes(x = ODBA, fill = individual.local.identifier))+
  facet_wrap(~individual.local.identifier)+
  ylim(0, 2500)

```
Pitch

```{r}
ggplot(odba_df)+
  geom_histogram(aes(x = pitch, fill = individual.local.identifier))+
  facet_wrap(~individual.local.identifier)+
  ylim(0, 2500)

```


